Predicting Canadian Grocery Price Trends Using Bayesian Models

### 2024-12-03 paper/paper.qmd
Updated paper

library(tidyverse)
library(brms)
library(ggplot2)
library(lubridate)
library(arrow)
library(knitr)
library(modelsummary)
library(rstan)

@@

Overview paragraph

@@

Estimand paragraph

The key estimand of this study is the effect of vendor choice, product category, and temporal trends on grocery pricing across Canada. Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing and the role that seasonal and economic shifts play. Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors.

@@

Results paragraph

The analysis reveals that vendor choice plays a significant role in determining grocery prices, with large national vendors generally offering lower prices compared to local or regional options. Furthermore, we observe a clear seasonal trend in grocery pricing, with prices typically rising in the winter months and dropping during certain sales periods. The model also indicates that certain product categories, particularly organic and plant-based foods, are subject to higher price volatility, while staple items such as grains and protein foods remain relatively stable.

Why it matters paragraph

@@

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data describes the data used in the analysis, @sec-model details the Bayesian model and assumptions, @sec-results presents the results, and @sec-discussion offers a discussion of the findings and suggests areas for future research.

@@

The "Canadian Grocery Price Data" dataset was sourced from Project Hammer [@grocery_data], and a statistical analysis was performed using the R programming language [@citeR]. The primary objective of this analysis was to clean, explore, and visualize the data to identify which grocery vendor currently and in the future offers the most cost-effective options for individuals adhering to the 2019 CFG. Various R libraries were utilized throughout the process. The tidyverse library [@tidyverse] provides a comprehensive set of tools for data manipulation, cleaning, and visualization, enabling efficient handling of datasets. The lubridate library [@lubridate] simplifies working with date-time objects while arrow [@arrow] is used to handle large datasets efficiently, especially for columnar formats like Parquet. For Bayesian modeling, rstanarm [@rstanarm] and brms [@brms] offer powerful tools for fitting and interpreting multilevel models while bayesplot [@bayesplot] aids in the graphical exploration of Bayesian model results, enhancing the interpretation of complex models. A summary table showcasing 10 sample observations from the cleaned dataset is presented in @tbl-grocery-data-summary. The table was divided into Part 1 and Part 2 for better readability, but the original dataset is formatted as one continuous table.

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

@@

- Another outcome variable is the overall vendor score as seen in @fig-overall-vendor-score, which takes into account multiple criteria to evaluate vendors. The highest weight is put on affordability which is measured by the average price per unit of groceries from each vendor. Some additional factors taken into account are whether products are organic (quality), are in stock (reliability), are on sale or are best sellers (appeal), and if there was a quiet price decrease (transparency).

@@

- Vendor Name (vendor): There are 8 vendors in this analysis: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria. This variable is crucial since we are comparing results between these vendors. For instance, we can factor in the fact that Voila has approximately 4 times the number of products compared to Galleria as seen in the table below.

@@

- Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by identifying instances where a product’s price decreased without being labeled as a sale. This outcome sheds light on subtle pricing strategies vendors might employ to remain competitive without overt promotional activities. It provides insights into whether vendors prioritize transparent discounting (sales) or discreet price reductions. As seen in @fig-price-decrease, Metro has the most number of products with decreased prices that were not on sale.

@@

- Organic Product Status (is_organic): Whether a product is “Organic” serves as a critical predictor for those who want higher quality products while risking affordability. We compare the availability of high-quality products across vendors and categories by looking for product names that include “fresh” or “organic”. In @fig-is-organic, we see that Metro and Voila have the widest range of organic products. These products tend to be more pricey than normal products, which is a trade-off for people who want to invest in better quality over affordability.

@@

### Model Justification

@@

## Implementation and Validation  
The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit.

@@

The average weekly meal plan cost by vendor and food category in @fig-avg-price-stacked shows in order which vendors overall are the most cost-effective to the least cost-effective. By looking at the different colors that represent the food catgory, it can also be compared which food group is more cost-effective in which store.

@@

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Paper Summary {#sec-first-point}
what was done in this paper?

@@

According to the predicted future price trends in @fig-future-price-trends, Galleria exhibits relatively stable predicted price trends, with slight variations over time. The predicted prices stay consistent, without significant spikes, suggesting steady pricing strategies. Loblaws shows stable prices with slight fluctuations but no dramatic increases or decreases. This implies consistent pricing practices with minor seasonal or product-related variations. Metro displays a similar pattern of stability to Loblaws but with marginally larger variations. The periodicity of the vertical bars may indicate uncertainty in the predictions during certain months. NoFrills has a noticeable spike mid-period, followed by a return to stability. This indicates a potential pricing anomaly or predicted fluctuation in a specific month before returning to normal pricing levels. SaveOnFoods presents consistent prices with some minor upward shifts, suggesting a gradual increase or variability in certain product categories. However, sporadic trends can be seen where there is an abrupt jump between months, which could reflect noise in the data. T&T has a slight peak mid-period, followed by stabilization. This may reflect a seasonal effect or a one-time price fluctuation for specific products. Voila shows a small spike mid-period but generally maintains consistent prices. The predicted pricing pattern mirrors a temporary fluctuation, likely tied to vendor or product-specific dynamics. Walmart demonstrates the most volatility in its price trends, with noticeable spikes. This may highlight significant uncertainty in Walmart's pricing patterns or irregularities caused by promotional events or external market forces. The vertical lines represent the uncertainty intervals (credible intervals) of the Bayesian predictions. Wider intervals imply greater uncertainty about the predicted prices for that month. These intervals can stem from limited data availability for specific vendors or products, variability in historical pricing trends, or external factors influencing the model’s predictive accuracy.

@@

@fig-trace-plot shows the traceplots for the bayesian model. [Add more here]

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace-plot
#| fig-cap: "Checking the convergence of the bayesian model using trace plot"
#| layout-ncol: 2

plot(bayesian_model)

>>>>

library(tidyverse)
library(brms)
library(ggplot2)
library(lubridate)
library(arrow)
library(knitr)
library(modelsummary)
library(rstan)
library(bayesplot)

@@

@@

The key estimand of this study is the effect of vendor choice, product category, and temporal trends on grocery pricing across Canada. Specifically, we aim to estimate how much more or less consumers might pay for certain product categories when purchasing from different vendors, considering the ongoing trends in grocery pricing. Through this analysis, we seek to estimate the probability of future price decreases or increases based on these factors.

@@

The analysis reveals that vendor choice plays a significant role in determining grocery prices when factoring product categories and temporal attributes. NoFrills offers the lowest predicted prices, making it ideal for budget-conscious shoppers, though it has occasional price spikes. For those valuing price stability, Loblaws or Metro are better options. However, Voila emerges as the top choice, balancing affordability with superior service and reliability.

@@

The remainder of this paper is structured as follows. @sec-data describes the data used in the analysis, @sec-model details the Bayesian model and assumptions, @sec-results presents the results, and @sec-discussion offers a discussion of the findings and suggests areas for future research. An overview of the sampling methdology is in @sec-appendix-A. Additional data details are explained in @sec-appendix-B, and additional model details are noted in @sec-appendix-C.

@@

The "Canadian Grocery Price Data" dataset was sourced from Project Hammer [@grocery_data], and a statistical analysis was performed using the R programming language [@citeR]. The primary objective of this analysis was to clean, explore, and visualize the data to identify which grocery vendor currently and in the future offers the most cost-effective options for individuals adhering to the 2019 CFG. Various R libraries were utilized throughout the process. The tidyverse library [@tidyverse] provides a comprehensive set of tools for data manipulation, cleaning, and visualization, enabling efficient handling of datasets. The lubridate library [@lubridate] simplifies working with date-time objects while arrow [@arrow] is used to handle large datasets efficiently, especially for formats in Parquet. For Bayesian modeling, rstanarm [@rstanarm] and brms [@brms] offer powerful tools for fitting and interpreting multilevel models while bayesplot [@bayesplot] aids in the graphical exploration of Bayesian model results, enhancing the interpretation of complex models. A summary table showcasing 10 sample observations from the cleaned dataset is presented in @tbl-grocery-data-summary. The table was divided into Part 1 and Part 2 for better readability, but the original dataset is formatted as one continuous table.

@@

@@

- Another outcome variable is the overall vendor score as seen in @fig-overall-vendor-score, which takes into account multiple criteria to evaluate vendors. The highest weight is put on affordability since it is the primary goal of this paper, and is measured by the average price per unit of groceries from each vendor. Some additional factors taken into account are whether products are organic (quality), are in stock (reliability), are on sale or are best sellers (appeal), and if there was a quiet price decrease (transparency). The last three factors had the lowest weights because it did not apply to every vendor.

@@

- Vendor Name (vendor): There are 8 vendors in this analysis as shown in the table below: Voila, Loblaws, NoFrills, Metro, Walmart, TandT, SaveonFoods, and Galleria. This variable is crucial since we are comparing results between these vendors.

@@

- Quiet Price Decrease (price_decrease): Tracks subtle pricing adjustments by identifying instances where a product’s price decreased without being labeled as a sale. This outcome indicates subtle pricing strategies vendors might employ to remain competitive without overt promotional activities. It provides insights into whether vendors prioritize transparent discounting (sales) or discreet price reductions. As seen in @fig-price-decrease, Metro has the most number of products with decreased prices that were not on sale.

@@

- Organic Product Status (is_organic): Whether a product is “Organic” serves as a predictor for those who want higher quality products while risking affordability. We compare the availability of high-quality products across vendors and categories by looking for product names that include “fresh” or “organic”. In @fig-is-organic, we see that Metro and Voila have the widest range of organic products. These products tend to be more pricey than normal products, which is a trade-off for people who want to invest in better quality over affordability.

@@

## Model Justification

@@

## Implementation and Validation  
The model was implemented in R using the `brms` package. Validation was done by visually checking posterior predictive checks to assess the model's fit and is explained with more detail in @sec-appendix-C.

@@

The average weekly meal plan cost by vendor and food category in @fig-avg-price-stacked ranks vendors from most to least cost-effective. The color-coded food categories also allow for comparisons of which food groups are more cost-effective at specific stores.

@@

## Paper Summary {#sec-first-point}
This paper explored several complementary analyses to evaluate grocery pricing trends across vendors and food categories. These efforts provided deeper insights into affordability and purchasing strategies aligned with the Canadian Food Guide recommendations.

Firstly, the average weekly meal plan cost analysis revealed significant differences among vendors, offering a detailed breakdown of pricing trends across food categories. This visualization highlighted which vendors were most cost-effective overall and how food group pricing varied between stores. The additional breakdown of costs by food group identified which categories—fruits and vegetables, protein, grains, or dairy—had the highest and lowest costs across vendors.

Vendor-specific attributes, such as sale transparency, product appeal, and reliability of being in stock, were also examined as part of a broader evaluation. While affordability remained a dominant factor, these additional features provided context for choosing vendors based on consumer priorities beyond cost. For example, stores that prioritize organic products or explicitly promote sales emerged as better choices for consumers seeking environmentally conscious or clearly discounted options.

This research also underscored the importance of informed meal planning, where a combination of affordability and quality factors can guide consumer decisions. The results have implications for public policy, particularly in addressing food accessibility and encouraging healthier eating habits by promoting cost-effective options for meeting dietary recommendations.

@@

According to the predicted future price trends in @fig-future-price-trends, most vendors show relatively stable predicted price trends, with slight variations over time. However, the periodicity of the vertical bars in the graphs may indicate uncertainty in the predictions during certain months. Galleria has a sharp decrease followed by a sharp increase halfway into the year, showing unstability in pricing trends. Loblaws and Metro have minor variations and seem overall more stable. NoFrills has a noticeable spike mid-period, followed by a return to stability. This indicates a potential pricing anomaly or predicted fluctuation in a specific month before returning to normal pricing levels. SaveOnFoods presents consistent prices with some minor upward shifts, suggesting a gradual increase or variability in certain product categories. However, sporadic trends can be seen where there is an abrupt jump between months, which could reflect noise in the data. T&T has a slight peak mid-period, followed by stabilization. This may reflect a seasonal effect or a one-time price fluctuation for specific products. Voila shows a small spike mid-period but generally maintains consistent prices. Walmart demonstrates the most volatility in its price trends, with noticeable spikes. This may highlight significant uncertainty in Walmart's pricing patterns or irregularities caused by promotional events or external market forces. The vertical lines represent the uncertainty intervals (credible intervals) of the Bayesian predictions. Wider intervals imply greater uncertainty about the predicted prices for that month. These intervals can stem from limited data availability for specific vendors or products, variability in historical pricing trends, or external factors influencing the model’s predictive accuracy.

@@

The traceplots for the bayesian model are shown in three separate figures, @fig-trace-plot-group1, @fig-trace-plot-group2, and @fig-trace-plot-group3 for better visibility. These plots display the sampled values for each parameter across iterations for two chains. Ideally, the chains should mix well and overlap to indicate convergence. Overall, the appearance of the trace plots suggest that the Bayesian model has converged well for all parameters used in the model. Both chains overlap and explore the posterior space effectively, with no significant signs of non-convergence. This ensures the use of these posterior estimates in the inference and analysis of this paper is appropriate.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false

posterior_samples <- as.matrix(bayesian_model)

# Group 1
group1_params <- c("b_Intercept", "b_current_price", "b_food_categoryGrainProducts", "b_food_categoryProteinFoods", "b_vendorLoblaws")
mcmc_trace_group1 <- mcmc_trace(posterior_samples[, group1_params]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Group 2
group2_params <- c("b_vendorMetro", "b_vendorNoFrills", "b_vendorSaveOnFoods", "b_vendorTandT", "b_vendorVoila")
mcmc_trace_group2 <- mcmc_trace(posterior_samples[, group2_params]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Group 3
group3_params <- c("b_vendorWalmart", "b_month", "b_price_decrease", "sd_product_name__Intercept", "sigma")
mcmc_trace_group3 <- mcmc_trace(posterior_samples[, group3_params]) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace-plot-group1
#| fig-cap: "Trace plots for first five predictors"

mcmc_trace_group1
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace-plot-group2
#| fig-cap: "Trace plots for sixth to tenth predictors"

mcmc_trace_group2
```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace-plot-group3
#| fig-cap: "Trace plots for last five predictors"

mcmc_trace_group3
```
